\textbf{Too Beautiful to be Fake: Attractive Faces are Less Likely to be Judged as Artificially Generated}

### 2022-09-13 submission/manuscript.Rmd
update intro

  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "LKC Medicine, Nanyang Technological University, Singapore"
  - id            : "3"
    institution   : "National Institute of Education, Singapore"
  - id            : "4"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"
authornote: |
  Correspondence concerning this article should be addressed to Dominique Makowski, HSS 04-18, 48 Nanyang Avenue, Singapore (dom.makowski@gmail.com).
abstract: |
  Abstract abstract abstract.
keywords          : "visual illusions, illusion game, Pyllusion, personality, general factor"
wordcount         : "5114"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
---

@@

<!-- Objective features -->
Technology allows already to create near-perfect simulations indistinguishable from reality. If real and fake stimuli are identical based on their objective characteristics, how can we make - hopefully accurate - judgments regarding their nature?

<!-- Context -->
Literatue shows that context surrounding stimuli plays an important role. But in the absence of context, what drives the beliefs of reality?

<!-- Internal features -->
Psychological factors. Stable dispositional traits. Reactive characteristics.

>>>>

  - id            : "1"
    institution   : "School of Social Sciences, Nanyang Technological University, Singapore"
  - id            : "2"
    institution   : "LKC Medicine, Nanyang Technological University, Singapore"
  - id            : "3"
    institution   : "National Institute of Education, Singapore"
  - id            : "4"
    institution   : "Centre for Research and Development in Learning, Nanyang Technological University, Singapore"
authornote: |
  Correspondence concerning this article should be addressed to Dominique Makowski, HSS 04-18, 48 Nanyang Avenue, Singapore (dom.makowski@gmail.com).
abstract: |
  Abstract abstract abstract.
keywords          : "visual illusions, illusion game, Pyllusion, personality, general factor"
wordcount         : "5114"
bibliography      : references.bib
floatsintext      : yes
linenumbers       : yes
draft             : no
mask              : no
figurelist        : yes
tablelist         : no
footnotelist      : no
classoption       : "man"
output            : papaja::apa6_pdf
csl: utils/apa.csl
header-includes:
  - \usepackage[labelfont=bf, font={scriptsize, color=gray}]{caption}
editor_options:
  chunk_output_type: console
bibliography: references.bib
---

@@

<!-- stakes and Post-truth era -->
For the first time in the history of humanity, technology has enabled the creation of near-perfect simulations  indistinguishable from reality. These artificial yet realistic constructs permeate all areas of life through immersive works of fiction, deep fakes (real-like images and videos generated by deep learning algorithms), virtual and augmented reality, artificial beings (artificial intelligence "bots" with or without a physical form), fake news and skewed narratives, of which ground truth is often hard to access [@nightingale2022]. Such progress not only carries important consequences for the technological and entertainment sectors, but also for security and politics, for instance if used for propaganda and  disinformation, recruitment into malevolent organizations, or religious indoctrination [@pantserev2020]. This challenge is  thus central to what has been coined the "post-truth era" [@lewandowsky2017beyond], in which the distinction (and lack  thereof) between authentic and simulated objects will play a critical role.

<!-- Ambiguous stimuli and irrelevance of objective features -->
While not all simulations have achieved perfect realism (e.g., Computer Generated Images - CGI in movies often lack certain  key details and an overall appearance of reality), it is fair to assume that these technical limitations will become  negligible in the near future. In fact, in the field of face perception and recognition, recent research found images of  artificially-generated faces to be perceived as more real than real faces [@nightingale2022; @tucciarelli2020; @moshel2022]. This fact, however, leads to a new issue: if real and fake stimuli cannot be distinguished based on their objective  characteristics (for instance, facial features in the perception of real relative to fake faces; [@holmes2016; @moshel2022]), how can we make judgments regarding their nature?

<!-- Context -->
Literature shows that the context surrounding a stimulus often plays an important  role in the assessment of its reality (a process henceforth referred to as *simulation monitoring* [@makowski2019phenomenal; @makowski2018cognitive]). Whereas individuals could rely on the content of the stimulus, making a reality judgement based solely on this information could be too cognitively effortful, especially if the stimulus is imbued with ambiguity [@michael2021source; @susmann2021persuasion]. As such, the Elaboration Likelihood Model(ELM) posits that people tend to draw on peripheral contextual cues, such as the source of the stimulus, credibility, authority and expertise of the source etc., to help facilitate their evaluation [petty1986elaboration]. Accordingly, such reliance on peripheral cues has been consistently reported across multiple studies, most recently with regards to people's trust in news articles reporting on the COVID-19 pandemic [@susmann2021persuasion;@michael2021source]. In the absence of these contextual information, however, what then  drives our beliefs of reality? (**Figure 1**).

```{r message=FALSE, warning=FALSE, fig.cap="The decision to believe that an ambiguous stimulus (of any form, e.g., images, text, videos, environments, ...) is real or fake depends of individual characteristics (e.g., personality and cognitive styles), stimulus-related features (context, emotionality), and their interaction, which can manifest for instance in our bodily reaction.", out.width="100%"}
knitr::include_graphics("../figures/Figure1.png")
```

<!-- Internal features -->
Similar to contextual cues, the inter-individual characteristics of the assessor also impacts whether a stimulus is ultimately perceived to be real (or fake). Factors such as the individual's cognitive style, prior beliefs, as well as personality traits have been shown to significantly impact people's belief in ambiguous stimuli [@bryanov2021determinants; @ecker2022psychological; @sindermann2020short]. Specifically, individuals with higher levels of analytical reasoning have been found to be better at distinguishing real from fake stimuli, possibly since detection of false information requires careful scrutinization and deliberation [@pehlivanoglu2021role; @pennycook2019lazy]. Unsurprisingly, possessing prior knowledge or beliefs about the stimulus at hand also influences one's perception of it; biasing us to attend only to information in line with our predispositions (what is known as **confirmation bias**) [@britt2019reasoned]. For instance, in a study conducted by [@tucciarelli2020], participants who were told about the presence of social agents reported lower levels of trust relative to those who were not relayed any information. Furthermore, dispositional traits, such as high levels of narcissism and low levels of openness and conscientiousness have been associated with greater susceptibility to fake news [@piksa2022cognitive; @sindermann2020short].

<!-- Reactive features -->
Aside from stimulus- and individual-related characteristics, it is possible that simulation monitoring is driven by an interaction between the two, i.e., by the reaction associated with the experience of a given stimulus. One way in which this is exemplified is via emotions. For instance, in the field of virtual reality (VR), emotional features have been shown to increase the virtual domain's presence, i.e., the extent to which the the system is able to deliver a convincing (real-like) environment with which an agent can interact [@sanchez2005presence; @makowski2017being]. This hence implies that the inclusion of emotional features could increase people's belief in simulations. Indeed, several studies have suggested that heightened stimulus emotionality is associated with greater belief in fake, but not real, information [@martel2020reliance]. On the contrary, research relating emotional works of fiction to simulation monitoring have instead found emotionality to be associated with reduced perceptions of realness [@makowski2019phenomenal], i.e., emotional stimuli were more likely perceived to be fake as opposed to real. However, since participants were told the descriptions corresponding to some of the stimuli were fake, the latter finding could be understood as a result of the participants engaging in spontaneous fictional appraisal, an emotion regulation mechanism in which individuals lower their perception of realness of the given stimulus by endowing it with a fictional context [@sperduti2017distinctive; @sperduti2016paradox]. Moreover, when the stimuli was self-relevant, studies show that the experienced intensity of emotions was enhanced [@sperduti2016paradox; @goldstein2009pleasure]. This hence implies that self-relevance could modulate the relationship between the emotionality of a stimulus and our perceptions of it's realness. Additionally, other features that drive individuals' evaluation of a stimulus, such as familiarity, have also been shown to correlate positively with the perception of it as truth [@begg1992dissociation]

<!-- Attractiveness -->
Images of faces, one of the most common artificial intelligence (AI) targets, integrate components of emotional reaction, saliency, self-relevance via attractiveness (REF). Talk about possible links between attractiveness and simulation monitoring

<!-- Goals and hypotheses -->

```{r message=FALSE, warning=FALSE, fig.cap="Top part shows blabla.", out.width="100%"}
knitr::include_graphics("../figures/Figure2.png")
```

## Methods

### Procedure

### Participants

### Data Analysis

The real-fake ratings were converted into two scores.

## Results

The analysis was carried out using *R 4.2* [**REF**], the *tidyverse* [**REF**], and the *easystats* collection of packages [**REFs for insight, parameters, performance, see**].

### Manipulation Check

Only one image file, classified as fake in 87.4\% of trials, yielded a strong simulation monitoring bias (> 85\%). This image was removed from further analysis (leaving 108 trials per participant). On average, across participants, 44\% of trials (95\% CI [0.11, 0.64]) was considered fake and 56\% of trials (95\% CI [0.36, 0.89]) was considered real.

There was a significant negative effect of the delay of re-exposure (with 95\% of values between 1.58 and 30.31 min), suggesting that shorter delays were associated with a slight bias towards the belief of reality (60\% at a theoretical delay of 0), which decreased to 50\% at a theoretical delay of 60 min ($OR = 0.99$, $95\% CI = [0.99, 1.00]$ $z = -2.27$, $p = 0.023$). There was also a significant negative effect on judgment confidence, but only in the real condition ($\beta = -0.005$, $95\% CI = [-0.1, 0.0]$, $p = .023$).

### Determinants of Simualtion Monitoring

## Discussion

Although expected, the first striking result is that despite all items being real photos from the same database, all people easily believe (to high degrees of confidence) that a significant proportion of them was fake. This is a testimony to the existing expectations regarding CGI technology, as much as to the volatility of our sense of reality.
